---
title: Notes On Everything
author: Desmond Tuiyot
date: '2020-05-24'
slug: notes-on-everything
categories:
  - Statistics
  - Personal Notes
tags:
  - Data Analysis
  - Notes
  - Random
bibliography: ref.bib
# csl: style.csl 
---

```{r include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```
## Disclaimer:  
These are my personal notes on things I learn. They are neither extensive nor comprehensive and will probably not be useful to you. There may be a lot of mistakes and errors because this space represents my attempts at understanding the things I learn.

## Statistics {#statistics}

## Probability {#probability}

## Machine Learning {#machine_learning}

## Algorithms & Data Structures {#algos_data_structures}
### Dynamic Programming - Patterns  
1. Minimum/Maximum path to reach a target  
These are straightforward optimization questions. You have a target to reach and at each step, you have some number of possible paths to take. The approach:
  + The subproblem is the minimum/maximum path to step `i`, where `0<=i<=n`
  + Choose the minimum/maximum path amongst all possible paths to reach the current state. Then add the value for the current state. 
  + Generate optimal solutions for all values leading up to and including the target, then return the value for the target. 
For each problem, I have to first think about the base cases, which I use to build the solution once I start iterating.  
The subproblem in this family of problems seem to be primarily prefixes - At subproblem `DP(i)`, I'm assuming that all the subproblems `DP(i)` depends on are already solved. 

## Python

## RShiny

### Inputs & Outputs
#### Inputs
textInput()  
sliderInput()  
selectInput()  
numericalInput()  
dateRangeInput()  
all inputs have the format
```{r echo=FALSE, eval=FALSE}
library(shiny)
library(dplyr)
library(tidyr)
```
``` {r eval=FALSE}
__input("inputId",
        "label",
        unique_param1, unique_param2, ...)
```
#### Render Functions
Render functions are used to built outputs in the server based on
inputs and possibly other stuff
```{r eval=FALSE}
renderText()
renderTable()
renderImage()
renderPlot()
```
#### Output Functions
Output functions are used in the ui to display the result built by 
render functions in the server
```{r eval=FALSE}
textOutput()
plotOutput()
tableOutput() or dataTableOutput()
imageOutput()
```
##### Non-shiny output and render functions
DT, leaflet, and plotly -> interactive data tables, maps, and plots as Shiny outputs  
In order to add an output to a Shiny app, we need to:  
1. Create the output -> could be a plot, table, string, etc
2. Render the output in the `server` function using appropriate `Render___` function.
3. Assign this render to a variable name prefixed with `output$___`
3. Use the corresponding `___Output` and pass in the variable name
### Layouts and Themes
#### Layouts
1. Sidebar layouts
```{r eval = FALSE}
sidebarLayout(
  sidebarPanel(insert input/output here),
  mainPanel(insert input/output here)
)
```
2. Tab layouts
```{r eval=FALSE}
sidebarLayout(
  sidebarPanel(insert input/output here),
  mainPanel(
    
    tabsetPanel(
      tabPanel(),
      tabPanel()
    )
  )
)
```
#### Themes
Add a theme selector to your Ui
```{r eval=FALSE}
shinythemes::themeSelector()
```
Then you can add it to your U  
### Building The App: A Process  
1. Add inputs to the `ui()`
2. Add outputs to the `ui()`/`server()`
3. Modify the app layout in the `ui()`
4. Update the output in the `server()` to incorporate the input  
## References {#references}
